Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-08 drafts/hybrid_zone.tex
minor fixes

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation. % too much `understanding' here
Here, we use a combination of analytical theory and explicit forward simulations
to show how selection against hybrid genotypes impacts patterns of introgression  % too much `understanding' here
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  physical scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location. 
Our findings suggest that the density of ancestry junctions provide a means for characterizing hybrid zones and identifying candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}
}

@@

%%%%%%%%
\begin{figure}
\includegraphics{figs/BM_schematic}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    i.e., showing where (horizontal axis) and when (vertical axis) the ancestors, from whom the haplotype was inherited, lived.
    The segment contains the $B$ alelle at the selected locus.
    The black line shows the lineage at the selected site in the sampled individual in the present day (at $\tau=0$).
    The center of the hybrid zone is at geographic position $x=0$; in the region $x>0$, ancestry $A$ is less common,
    and because of negative selection, the selected lineage spends little time there.
    Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
    these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
    or can switch lineages at linked sites between selected backgrounds 
    (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
    Here the initial haplotype is broken up into 3 chunks over 100 generations,
    which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations after contact, 
    from left to right the segments are of ancestry $(A,B,B)$.
}\label{Fig:schematic}
\end{figure}

@@

\begin{figure}
\includegraphics{figs/alleleFrequencies_sim}
\caption{
    Frequency of ancestry $B$,
    across geography at different physical positions on the genome, simulated for a hybrid zone
     $T=100$ generations after secondary contact, with $s=0.1$,
     using 50 demes, each with 500 diploid individuals and $\sigma=1$.
    Each line represents a locus some distance $r$ away from the true target of selection, 
    with $r=0$ representing the locus under selection. 
    Grey lines represent the same positions from a simulation with
    identical parameters except that $s=0$.
    Corresponding theoretical quantities are shown juxtaposed in Figure \ref{sfig:alleleFreq_tau100_comparison};
    the same plot is shown with weaker selection in Figure \ref{alleleFreq_tau100_weaker_s} and at a longer time in Figure \ref{alleleFreq_tau1000}.
}\label{alleleFreq_tau100}
\end{figure}

@@

\paragraph{Identifying selected loci}
The statistic $C(m,x)$ shows promise for identifying selected loci under some circumstances.
As expected, regions surrounding a locus under selection are more resistant to introgression,
as seen in Figures~\ref{Fig:resistanceToIntrogression1000g} and \ref{Fig:resistanceToIntrogression100g}).
When present, we expect haplotypes that contain the locally less common allele
to be longer than the genome-wide average.
Indeed,
as shown in Figures \ref{Fig:blockLengths}, \ref{Fig:blockLengthsZoom} and \ref{Supp:blockLengthHeatmap},
the mean length of such haplotype blocks is up to three times longer
than the average for that geographic location,
peaking quite sharply around the location of the selected site.

Genomic regions are inherited from ancestors $T$ generations ago
in blocks of size roughly $1/T$.
There will be a long segment of $A$ ancestry
if many such adjacent blocks are all inherited from the $A$ side.
Ancestries of neighboring blocks are correlated,
due to the branching process described above. 
But, if we assume they are independent, 
then since the block surrounding location $r$ in an individual at geographic location $x$ is of ancestry $A$ with probability $q_A(x,T,r)$,
we'd expect to see, roughly, 
$q_A(x,T,r)(1+2q_A(x,T,r))/(1-q_A(x,T,r))$ 
% P(N=0) = q and if N>0 then is 1 + two independent q-geometrics with mean q/(1-q),
% so is q ( 1 + 2q/(1-q) ) = q(1+2q)/(1-q) .
consecutive blocks of ancestry $A$ about a given site unlinked to the selected site.
(This assumes, crudely, that the number of $A$-blocks on either side of the enclosing block
has a Geometric distribution with parameter $q_A(x,T,r)$.)
This implies the mean length $l(x,r)$ would be $q_A(x,T,r)(1+2q_A(x,T,r))/((1-q_A(x,T,r))T)$,
and the mean conditional length $l_A(x,r)$ would be $(1+2q_A(x,T,r))/(T(1-q_A(x,T,r))$.
Furthermore, we know that if $x=0$ or if $T$ is large and $r$ is not too small, that $q_A$ is close to $1/2$
(so $l_A(x,r) \approx 4/T$);
and for $r<1/T$ that $q_A$ looks like the selected cline.
Also, we know from the discussion above that the lineage of a \emph{selected} $A$ allele,
if it is in the region where $A$ is rare,
moves at speed roughly $\sigma \sqrt{s}$ back towards the $A$ side of the zone,
returning to the region where $A$ is common in about $x/(\sigma \sqrt{s})$ generations.
Therefore, a selected $A$ allele found on the $B$ side of the zone
should carry with it a haplotype of average length $\sigma \sqrt{s}/x$
that looks like haplotypes from the center of the zone.
This analysis suggests that $A$ haplotypes in the center of the zone should be of average length $4/T$;
this is indeed what is seen at distant sites, for instance, in Figure \ref{sfig:blocksAlongChromAncBConditioning_nonnormalized}.
Haplotypes at the selected site are expected to be longer, but still of a length proportional to $1/T$,
suggesting that the normalization in the statistic $C(x,m)$ is appropriate,
as shown in Figure \ref{Supp:adjacentBlocks},
although a numerical prediction of the value of $C(x,m)$ is elusive.
\alisa{Above argument assumes independence between lineages. Does this make block lengths conservative? }

>>>>

\abstract{
Hybrid zones formed between recently diverged populations offer an opportunity 
to study the mechanisms underlying reproductive isolation and the process of speciation.
Here, we use a combination of analytical theory and explicit forward simulations
to show how selection against hybrid genotypes impacts patterns of introgression
across genomic and geographic space. 
By understanding how lineages move across the hybrid zone, in a model without coalescence, 
we describe how clines form and how parental haplotypes are broken up during introgression.
Clines form in about $1/s$ generations, where $s$ is the strength of selection against hybrids,
and linked clines persist over a  physical scale of $1/T$, where $T$ is the age, in generations, of the hybrid zone. 
Locally disadvantageous alleles tend to exist as small families,
whose lineages trace back to the side from which they originated at speed $\sqrt{s}$ dispersal distances per generation.
The lengths of continuous tracts of ancestry provide an additional source of information:
blocks of ancestry surrounding single-locus incompatibilities can be substantially longer than the genome-wide average block length at the same spatial location,
an observation that might be used to characterize the age of hybrid zones
and identify candidate targets of selection.
%\alisa{ascertainment bias. Tract lengths provide additional source of information, as they provide unbiased information even when used ascertained markers.}
}

@@

%%%%%%%%
\begin{figure}
\includegraphics{figs/BM_schematic}
\caption{
    \textbf{A schematic figure of the lineages along which a segment of genome has been inherited,}
    showing the location (horizontal axis) and time (vertical axis) at which the ancestors 
    of a hypothetical sampled haplotype lived.
    The segment contains the $B$ alelle at the selected locus.
    The black line shows the lineage at the selected site in the sampled individual in the present day (at $\tau=0$).
    The center of the hybrid zone is at geographic position $x=0$; in the region $x<0$, ancestry $B$ is less common,
    and because of negative selection, the selected lineage spends little time there.
    Red dots represent ancestral recombination events that fall in the given segment of genome in an ancestor;
    these can cause the lineage to branch (lineages at linked sites are depicted as colored lines),
    or can switch lineages at linked sites between selected backgrounds 
    (yellow lineages are linked to the $A$ selected locus; blue are linked to $B$).
    Here the initial haplotype is broken up into 3 chunks over 100 generations,
    which switch between selected backgrounds once.
    Based on position of the lineages at $\tau=100$ generations ago,
    from left to right the segments are of ancestry $(A,B,B)$.
}\label{Fig:schematic}
\end{figure}

@@

\begin{figure}
\includegraphics{figs/alleleFrequencies_sim}
\caption{
    Frequency of ancestry $B$,
    across geography at different physical positions on the genome, simulated for a hybrid zone
     $T=100$ generations after secondary contact, with $s=0.1$,
     using 50 demes, each with 500 diploid individuals and $\sigma=1$.
    Each line represents a locus some distance $r$ away from the true target of selection.
    Grey lines represent the same positions from a simulation with
    identical parameters except that $s=0$.
    Corresponding theoretical quantities are shown juxtaposed in Figure \ref{sfig:alleleFreq_tau100_comparison};
    the same plot is shown with weaker selection in Figure \ref{alleleFreq_tau100_weaker_s} and at a longer time in Figure \ref{alleleFreq_tau1000}.
}\label{alleleFreq_tau100}
\end{figure}

@@

\paragraph{Identifying selected loci}
The statistic $C(m,x)$ shows promise for identifying selected loci under some circumstances.
As expected, regions surrounding a locus under selection are more resistant to introgression,
as seen in Figures~\ref{Fig:resistanceToIntrogression1000g} and \ref{Fig:resistanceToIntrogression100g}.
When present, we expect haplotypes that contain the locally less common allele
to be longer than the genome-wide average.
Indeed,
as shown in Figures \ref{Fig:blockLengthsZoom}, \ref{Fig:blockLengths} and \ref{Supp:blockLengthHeatmap},
the mean length of such haplotype blocks is up to three times longer
than the average for that geographic location,
peaking quite sharply around the location of the selected site.

Genomic regions are inherited from ancestors $T$ generations ago
in blocks of size roughly $1/T$.
There will be a long segment of $A$ ancestry
if many such adjacent blocks are all inherited from the $A$ side.
Ancestries of neighboring blocks are correlated,
due to the branching process described above. 
But, if we assume they are independent, 
then since the block surrounding location $r$ in an individual at geographic location $x$ is of ancestry $A$ with probability $q_A(x,T,r)$,
we'd expect to see, roughly, 
$q_A(x,T,r)(1+2q_A(x,T,r))/(1-q_A(x,T,r))$ 
% P(N=0) = q and if N>0 then is 1 + two independent q-geometrics with mean q/(1-q),
% so is q ( 1 + 2q/(1-q) ) = q(1+2q)/(1-q) .
consecutive blocks of ancestry $A$ about a given site unlinked to the selected site.
(This assumes, crudely, that the number of $A$-blocks on either side of the enclosing block
has a Geometric distribution with parameter $q_A(x,T,r)$.)
This implies the mean length $l(x,r)$ would be $q_A(x,T,r)(1+2q_A(x,T,r))/((1-q_A(x,T,r))T)$,
and the mean conditional length $l_A(x,r)$ would be $(1+2q_A(x,T,r))/(T(1-q_A(x,T,r))$.
Furthermore, we know that if $x=0$ or if $T$ is large and $r$ is not too small, that $q_A$ is close to $1/2$
(so $l_A(x,r) \approx 4/T$);
and for $r<1/T$ that $q_A$ looks like the selected cline.
Also, we know from the discussion above that the lineage of a \emph{selected} $A$ allele,
if it is in the region where $A$ is rare,
moves at speed roughly $\sigma \sqrt{s}$ back towards the $A$ side of the zone,
returning to the region where $A$ is common in about $x/(\sigma \sqrt{s})$ generations.
Therefore, a selected $A$ allele found on the $B$ side of the zone
should carry with it a haplotype of average length $\sigma \sqrt{s}/x$
that looks like haplotypes from the center of the zone.
This analysis suggests that $A$ haplotypes in the center of the zone should be of average length $4/T$;
this is indeed what is seen at distant sites, for instance, in Figure \ref{sfig:blocksAlongChromAncBConditioning_nonnormalized}.
Haplotypes at the selected site are expected to be longer, but still of a length proportional to $1/T$,
suggesting that the normalization in the statistic $C(x,m)$ is appropriate,
as shown in Figure \ref{Supp:adjacentBlocks},
although a numerical prediction of the value of $C(x,m)$ is elusive.
% \alisa{Above argument assumes independence between lineages. Does this make block lengths conservative? }
% PR: yes, maybe, but that correction seems like a small one, and it's not getting very quantitative predictions;
% a more serious issue is that it's ignoring that the next block has a different $r$.
